%-----------------------------------------------------------------------------%
%- Greatest Common Divisors and Least Common Multiples -----------------------%
%-----------------------------------------------------------------------------%
\chapter{Greatest Common Divisors and Least Common Multiples}
\label{chap:GreatestcommonDivisorAndLeastCommoonMultiples}

\section{Divisibility}
\label{sec:Divisibility}
We say that ``$a$ divides $b$'' or ``$b$ is divisible by $a$'' if $b$ is a
multiple of $a$. The symbol to represent this we use a vertical bar.
\begin{align}
   a|&b  &~ \text{``a divides b''} & \\
   5∤&11 &~ \text{``5 does not divide 11''} & \\
  27|&0  &~ \text{``27 divides 0. (Think about $\frac{0}{27}$)''} & \\
   0|&0  &~ \text{``0 divides 0''} & \\
   0∤&6
\end{align}
Amongst the natural numbers $\mathbb{N} = \{0, 1, 2, 3, \ldots\}$, 0, 1 are
special. For the others, they are either prime or composite. \emph{Primes} are
divisible by exactly 2 numbers, themselves and 1. \emph{Composittes} are
divisible by numbers other than 1 and themselves.\footnote{1 is neither
composite nor prime. It is called \emph{the unit}.}

Example:\\
$\{2, 5, 7, 11, 13, 17, 19, 23 \}$ are all prime numers
$\{0,4,6,8,9,10,12,14,15,16,18,20,21,22\}$ are all composite numbers\\
\\
How many prime numbers are there? Infinitely many.\\
Proof:
\begin{align}
  \text{Suppose there were only finitely many primes:} \\
  S = \{P_1, P_2, P_3, \ldots P_n \} \\
  \text{Consider} \\
  P_1 \times P_2 \times P_3 \ldots \times P_n + 1 & = \ldots \\
   & = \ni S \\
  \text{Is not in S, so it cannot be a prime. So it must have a prime factor.
  But, none of the primes divide it, because they all leave a remainder of 1.
  So it cannot be composite either.} \\
  & \therefore \text{a contradiction.} \\
  & \Rightarrow \text{the set of primes must be infinite.}
\end{align}

The probability of a large number being prime is approximately $\frac{1}{log_n}$.

\section{Common Factors \& Greatest Common Factor}
What are the common factors of $28, 164$? \\
\\
Factors of $28 = \{ 1,2,4,7,14,28 \}$ \\
Factors of $164 = \{1, 2, 4, 41, 82, 164 \}$ \\
\\
The common factors are $\{1,2,4\}$ and the greatest common factor is $4$. \\
\\
What is the greatest common factor of $1234$ and $5762$?\\
\\
Doing many divisions is computationally costly and time consuming. We need a
better way to make this computation feasible.\\
\\
By using prime factors (ie factors of a composite number which are prime), we
can determine composite factors much faster:
\begin{align}
  164 & = 2 * 82 \\
      & = 2*2*41 \\
      & = 2^2 * 41 \\
\end{align} 
or
\begin{align}
  2^0 * 41^0 & = 1 \\
  2^1 * 41^0 & = 2 \\
  2^2 * 41^0 & = 4 \\
  2^0 * 41^1 & = 41 \\
  2^1 * 41^1 & = 82 \\
  2^2 * 41^1 & = 164
\end{align}
So $3^2 * 5^3 * 7^4$ has $3*4*5 = 60$ facotrs\\
\\
So factoring both numbers into products of primes will let us find the factors
and common factors easily and so might be a good way to go.\\
\\
GCF of $3^2 * 11^4 * 17^3$ is $8! = 1*2*3*4*5*6*7$ is easily worked out:
\begin{align}
  8! & = 2* 3 * 2^2 * 5 * 7 \\
     & = 2^4 * 3^2 * 5 * 7
\end{align}
So the greatest common factor is $3^2 = 9$.\\
\\
GCF of $2^5 * 3^2 * 11^4 * 17^3$ is $8! = 1*2*3*4*5*6*7$ is easily worked out:
\begin{align}
  8! & = 2* 3 * 2^2 * 5 * *2 * 3* 7 \\
     & = 2^4 * 3^2 * 5 * 7
\end{align}
So the greatest common factor is $2^4 * 3^2 = 9 * 18 = 144$.\\
\\
However, this is still computationally unfeasable for large numbers as it is
prohibitively time consuming so this method is also impractical, but, the
ancient Greeks knew how to do it easily.\footnote{This ancient Greek dude called
Euclid wrote a computer program in 300BC, but didn't have a PC to run it on so
he used a sandbox.}

\subsection{Recap}
If we know the prime factorization of $x$ and $y$ it is easy to find the
greatest common divisor.
\\
For example:
Calculate the GCD of $11!$
\begin{align}
     11! & = 1 * 2 * 3 * \ldots * 10 * 11 \\
  \text{or\ldots} & \\
         & = 2^8 * 3^4 * 5^2 * 7 * 11 \\
  (20/6) & = \frac{20* 19 * 18 * 17 * 16 * 15}{6 * 5* 4 * 3 * 2 * 1} \\
         & = 19 * 17 * 2^3 * 5 = 2^3 * 3 * 5 \\
  \therefore \text{GCD}(11!, (20/6)) &= 2^3 * 3 * 5 \\
         & = 120 \\
  \text{and} \text{LCM}(11!, (20/6)) & = 2^8 * 3^4 * 5^2 * 7 * 11 * 17 * 19 \\
\end{align}

\section{Euclid's Algorithm}
\label{sec:EuclidsAlgorithm}
\begin{enumerate}
  \item Take the biggest number, and find how many times the smallest number
  goes into it
  \item take the remainder
  \item Take smaller of your two numbers, and find how many times the
  remainder goes into the smaller number.
  \item Repeat previous two steps until you have a zero remainder; and the GCD
  is the smallest non-zero remainder previous to the zero.
\end{enumerate}
The last non-zero remainder is the greatest
\\
To Find the GCD of (123, 57):
\begin{align}
  123 & = 2 * 57 + 9 \\
   57 & = 6 * 9 + 3 \\
    3 & = 3 * 3 + 0 \text{look! no remainder!}
\end{align}
Note: If we reverse it, we can see that the GCD is a common factor of subsequent
remainders (3 is a multiple of 3, 6, and 9, as well as our original two
numbers, 123 and 57).
\\
To find the GCD of (11253, 3271):
\begin{align}
  11253 & = 3 * 3271 + 1440 \\
   3271 & = 2 * 1440 + 391 \\
   1440 & = 3 * 391 + 267 \\
    391 & = 1 * 267 + 124 \\
    267 & = 2 * 124 + 19 \\
    124 & = 6 * 19 + 10 \\
     19 & = 1 * 10 + 9 \\
     10 & = 1 * 9 + 1 \\
      9 & = 1 * 9 + 0 \\
      \therefore \text{GCD(11253,3271)} & = 1 \\
      \therefore \text{No common factors}
\end{align}
It appears to have no real use or importance at first, but is very
useful in things like cryptography.\footnote{How often does Bruce Schneier come
up to you and ask you for the GCD between two numbers? Simon Singh has asked me
only once, in November 2010.}
\\
If $a, b \in \mathbb{Z}$ and $g$ is the GCD, then there are integers $x$ and $y$
such that $ax + by = g$. Moreover, the equation $ax + by = c$ has solutions
$mathbb{Z} \in x,y$ iff $g|c$

Example:
\begin{align}
  123x + 57y & = 28 \\
  \text{where x and y have solutions in integers}
\end{align}